pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-08-04 bmc_article.tex
Updates from Overleaf

Since ARI is dominated by differences in the number of clusters (Additional File 1: Fig S2-3) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}). MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Additional File 1: Fig S3). Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters. We also tracked the mean F1 score and the ARI at the true number of clusters.

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Additional File 1: Fig S2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters. Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Additional File 1: Fig S28), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters. This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B). We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy. It nevertheless remained considerably slower than \textit{Seurat}. For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.

@@

\begin{enumerate}
   \item Filtering
   \begin{itemize}
     \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions - see for instance our \textit{default} filters in the methods) to exclude poor-quality cells while avoiding bias against some subpopulations.
     \item Features filtering based on feature type did not appear beneficial.
   \end{itemize}
   \item Normalization and scaling
   \begin{itemize}
     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
     \item \textit{sctransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
     \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy. We therefore advise against this practice.
   \end{itemize}
   \item Feature selection
   \begin{itemize}
     \item Deviance \cite{townesGlmpca2019} offered the best ranking of genes for feature selection.
     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features in our datasets.
   \end{itemize}
   \item Denoising/imputation
   \begin{itemize}
       \item Denoising appeared beneficial to the identification of some subpopulations in 10x datasets, but not in Smart-seq datasets.
       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances, although none of these methods was beneficial over all subpopulations.
   \end{itemize}
   \item PCA
   \begin{itemize}
     \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the 10 or 20 nearest neighbors).
   \end{itemize}
   \item Clustering
   \begin{itemize}
    \item We found \textit{scran}-based \textit{walktrap} clustering to show good performances, at however increased computational costs which might be prohibitive with large datasets.
    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates.
   \end{itemize}
\end{enumerate}

@@

All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here. scVI results and results of the clustering section were produced with \textit{pipeComp} 0.99.4, while all other results were produced with version 0.99.2 (\textit{biorxiv}) branch of the repository). All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation. All wrappers for methods used in the study can be found in the external data of the \textit{pipeComp} package (\url{https://github.com/plger/pipeComp/tree/master/inst/extdata}).

The gene counts for the two mixology datasets were downloaded from the  \href{https://github.com/LuyiTian/CellBench_data/data/sincell_with_class.RData}{CellBench repository}, commit 74fe79e. For the other datasets, we used the unfiltered counts from \cite{duoClustering2018}, available on the corresponding repository \url{https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison}. For simplicity, all starting \textit{SingleCellExperiment} objects with standardized metadata are available on \url{https://doi.org/10.6084/m9.figshare.11787210.v1}.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth, keepaspectratio]{{anova.pdf}}
    \caption{\textbf{Variance in clustering metrics explained by the parameters.} See methods for more detail.}
    \label{fig:overall}
\end{figure}

@@

\begin{table}[h!]
\caption{\textbf{Table S1: Overview of the benchmark datasets.} \# features' indicates the number of features detected in at least 10 cells.}
\label{tab:table1}
\begin{tabular}{rlrrl}
  \hline
dataset & {\color{red}\# features} & \# cells & protocol & description \\ 
  \hline
Koh & 33922 & 531 & SMARTer & 9 FACS purified differentiation stages \\ 
  Kumar & 41930 & 246 & SMARTer & Mouse ESC cultured in 3 different conditions \\ 
  Zhengmix4eq & 10434 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
  Zhengmix4uneq & 11369 & 6498 & 10x & Mixtures of FACS purified PBMCs \\ 
  Zhengmix8eq & 10600 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
  mixology10x3cl & 16208 & 902 & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
  mixology10x5cl & 11786 & 3918 & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
  simMix1 & 3696 & 2500 & 10x-based & Simulation of 10 human cell subpopulations \\
  simMix2 & 8893 & 3000 & 10x-based & Simulation of 9 mouse cell subpopulations \\
   \hline
\end{tabular}
\end{table}

@@

\section*{Additional Files}
  \subsection*{\textbf{Additional file 1}}
  Additional File 1: Fig S(PDF 16 MB)
  \subsection*{\textbf{Additional file 2}}
  More detailed description of \textit{pipeComp} and comparison with other tools. (PDF 0.5 MB)
\end{backmatter}
\end{document}

>>>>

Since ARI is dominated by differences in the number of clusters (Additional File 1: Fig S2-3) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}). MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Additional File 1: Fig S3). Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters, and we also tracked the ARI at the true number of clusters.

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}A, left) over \textit{Seurat}, although when used in combination with some networks it sometimes yielded a lower ARI or minimum precision. The ARI score at the true number of clusters, when available, showed similar performances, especially when using \textit{sctransform}. Because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Additional File 1: Fig S2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters. Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Additional File 1: Fig S28), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters. This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally comparable to \textit{Seurat} clustering, although the latter offered more control over the resolution. A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B). We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy. It nevertheless remained considerably slower than \textit{Seurat}. For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.

@@

\begin{enumerate}
   \item Filtering
   \begin{itemize}
     \item Doublet detection and removal is advised and can be performed at little computing cost with software such as \textit{scDblFinder} or \textit{scds}.
     \item Distribution-based cell filtering fails to capture doublets and should use relatively lenient cutoffs (e.g., 5 MADs, or 3 MADs in at least 2 distributions - see for instance our \textit{default} filters in the methods) to exclude poor-quality cells while avoiding bias against some subpopulations.
     \item Features filtering based on feature type did not appear beneficial.
   \end{itemize}
   \item Normalization and scaling
   \begin{itemize}
     \item Most normalization methods tested yielded a fair performance, especially when combined with scaling, which tended to have a positive impact on clustering.
     \item \textit{sctransform} offered the best overall performance in terms of the separability of the subpopulations, as well as removing the effect of library size and detection rate.
     \item The common practice of regressing out cell covariates, such as the detection rate or proportion of mitochondrial reads nearly always had a negative impact, leading to increased correlation with covariates and decreased clustering accuracy. We therefore advise against this practice.
   \end{itemize}
   \item Feature selection
   \begin{itemize}
     \item Deviance \cite{townesGlmpca2019} offered the best ranking of genes for feature selection.
     \item Increasing the number of features included tended to lead to better classifications, plateauing from 4000 features in our datasets.
   \end{itemize}
   \item Denoising/imputation
   \begin{itemize}
       \item Denoising appeared beneficial to the identification of some subpopulations in 10x datasets, but not in Smart-seq datasets.
       \item We found especially \textit{ALRA} (with prior normalization), \textit{DrImpute} (with prior processing) and \textit{DCA} to offer the best performances, although none of these methods was beneficial over all subpopulations.
   \end{itemize}
   \item PCA
   \begin{itemize}
     \item Similarly to previous reports \cite{SunDimRed2019}, we recommend the irlba-based PCA using weighting of the components, as implemented in \textit{Seurat}. 
     \item We advise against the commonly-used elbow method (because it is too conservative) and the jackstraw method (low performance at great computational costs) for deciding on the components to include, and recommend a method like the \textit{global maximum likelihood based on translated Poisson Mixture Model} method (e.g., implemented in the  \texttt{maxLikGlobalDimEst} function of \texttt{intrinsicDimension}, in our case using the 10 or 20 nearest neighbors).
   \end{itemize}
   \item Clustering
   \begin{itemize}
    \item In cases where prior knowledge can guide the choice of a resolution, \textit{Seurat} can be useful in affording manual control of it while, in the absence of such knowledge, \textit{scran}-based \textit{walktrap} clustering provided reasonable estimates, although at greater computational costs. 
   \end{itemize}
\end{enumerate}

@@

All analyses were performed through the \href{https://github.com/plger/pipeComp}{pipeComp} R package, which implements the pipeline framework described here. scVI results and results of the clustering section were produced with \textit{pipeComp} 0.99.4, while all other results were produced with version 0.99.2 (\textit{biorxiv}) branch of the repository). All code to reproduce the simulations and figures is available in the \url{https://github.com/markrobinsonuzh/scRNA\_pipelines\_paper} repository, which also includes the basic datasets with a standardized annotation. All wrappers for methods used in the study can be found in the external data of the \textit{pipeComp} package (\url{https://github.com/plger/pipeComp/tree/master/inst/extdata}). In addition, both the package and attending code are archived on \cite{figshare_code}.

The gene counts for the two mixology datasets were downloaded from the  \href{https://github.com/LuyiTian/CellBench_data/data/sincell_with_class.RData}{CellBench repository}, commit 74fe79e. For the other datasets, we used the unfiltered counts from \cite{duoClustering2018}, available on the corresponding repository \url{https://github.com/markrobinsonuzh/scRNAseq\_clustering\_comparison}. For simplicity, all starting \textit{SingleCellExperiment} objects with standardized metadata are available on \cite{figshare_sce}.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.55\textwidth, keepaspectratio]{{anova.pdf}}
    \caption{\textbf{Variance in clustering metrics explained by the parameters.} See methods for more detail.}
    \label{fig:overall}
\end{figure}

@@

\begin{table}[h!]
\caption{\textbf{Table S1: Overview of the benchmark datasets.} \# features' indicates the number of features detected in at least 10 cells.}
\label{tab:table1}
\begin{tabular}{rlrrl}
  \hline
dataset & \# features & \# cells & protocol & description \\ 
  \hline
Koh & 33922 & 531 & SMARTer & 9 FACS purified differentiation stages \\ 
  Kumar & 41930 & 246 & SMARTer & Mouse ESC cultured in 3 different conditions \\ 
  Zhengmix4eq & 10434 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
  Zhengmix4uneq & 11369 & 6498 & 10x & Mixtures of FACS purified PBMCs \\ 
  Zhengmix8eq & 10600 & 3994 & 10x & Mixtures of FACS purified PBMCs \\ 
  mixology10x3cl & 16208 & 902 & 10x & Mixture of 3  cancer cell lines from CellBench \\ 
  mixology10x5cl & 11786 & 3918 & 10x & Mixture of 5 cancer cell lines from CellBench \\ 
  simMix1 & 3696 & 2500 & 10x-based & Simulation of 10 human cell subpopulations \\
  simMix2 & 8893 & 3000 & 10x-based & Simulation of 9 mouse cell subpopulations \\
   \hline
\end{tabular}
\end{table}

@@

\section*{Additional Files}
  \subsection*{\textbf{Additional file 1}}
  Supplementary Figures (PDF 16 MB)
  \subsection*{\textbf{Additional file 2}}
  More detailed description of \textit{pipeComp} and comparison with other tools. (PDF 0.5 MB)
\end{backmatter}
\end{document}
